<!--
% \VignetteIndexEntry{The ExperimentHubData Package}
% \VignetteDepends{ExperimentHubData}
% \VignetteEngine{knitr::knitr}
-->

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

# The ExperimentHubData Package

**Package**: `r Biocpkg("ExperimentHubData")`<br />
**Authors**: `r packageDescription("ExperimentHubData")[["Author"]] `<br />
**Modified**: 24 December, 2015<br />
**Compiled**: `r date()`

## Overview

The ExperimentHubData package provides tools to add data resources to
Bioconductor's ExperimentHub. This 'hub' houses curated data from courses,
publications or experiments. Because these resources are the static endpoint of
a specific activity they are not expected to change over time (unless a bug is
found).  These resources are not raw files (as offered in `AnnotationHub`)
but instead `R` objects such as GRanges, SummarizedExperiment, data.frame etc.
Each resource has associated metadata that can be searched through the
`ExperimentHub` client interface.

This document describes how to add prepare and add resources to `ExperimentHub`. 

## Data package 

The data to be added must have an associated data package, similar in structure
to the `Bioconductor` experimental data packages. The key components of the
package are as follows:

* inst/scripts/
  - make-data script
  - get-data script

* inst/extadata/
 - metadata.csv

* vignettes/

* man/

* R/


### Data objects

#### Preparation

Data objects should be created with the script in inst/scripts/make-data.R.

#### Location

Data should not be included in the package but instead be made available in
svn, git or on an ftp server. The 'get-data' script in inst/scripts/ should
download the data to the local directory where the script is invoked.

### Test metadata

The `makeMetadataFromCsv` function in `ExperimentHubData` generates a list of
`ExperimentHubMetadata` objects. Test that the metadata.csv file in your
package is valid and read-able by `ExperimentHubData` by invoking the function
on your package.

```{r, makeMetadataFromCsv, eval=FALSE}
library(ExperimentHubData)
meta <- makeMetadtaFromCsv(pathToPackage="path/to/my/package")
```

Inspect the returned list and confirm all metadata fields are as expected.


## Storage of metadata

The metadata for a resource is stored in a MySQL database. When a user runs the
client the database is downloaded as a sqlite dump (cached copy) of the parent
MySQL backend. The combination of two databases are used because MySQL is more
performant when adding new records and the portability of sqlite is better for
local cached access.

## Adding resources

Once the data package is in order and the get-data script successfully
retrieves the data objects we are ready to add resources.

The `addResources` function performs two primary steps: push
data objects to AWS S3 buckets and insert metadata in the MySQL db. 

**TODO** set POST option
 
'metadataOnly = TRUE' will generate metadata but not download the data or push
to S3 buckets. 'insert = FALSE' will generate the metadata but not insert it in
the db. This is a good option for testing the metadata and confirming that all
looks ok.

```{r, eval=FALSE}
addResources(pathToPackage, metadataOnly = TRUE, insert = FALSE)
```

Downloading and pushing to S3 requires that aws command line tools is installed
installed and that the local directory has adequate disk space for the files.

Confirm you have aws command line tools installed and access help:

```{}
aws s3 help
```

Inserting metadata in the MySQL db requires access to the `ExperimentHub`
server. A convenient way to interact with the db is with the
ExperimentHubServer Docker container. This service downloads a copy of the
production db and enables local inserts and modifications. The changes can be
tested locally then dumped and copied back to the production db. Instructions
for using the ExperimentHubServer Docker can be found [here](**TODO**).
